---
title: "Colorectal Cancer (CRC) Study"
author: "Soham Ghosh"
date: "2024-06-18"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## CRC Data Analysis 

We applied Zinck to meta-analyze five metagenomic studies of CRC. The five studies correspond to five different countries for the CRC data, which are named "AT" (Australia), "US" (USA), "CN" (China), "DE" (Germany), and "FR" (France).  The sample sizes are 109, 127, 120, 114, and 104, respectively, and the number of cases and controls is roughly balanced in each study. We focus on all the species whose relative abundances are more than 0.2.

```{r CRC-data}
##################### CRC data species level ##########################
#######################################################################
library(randomForest)
library(zinck)
library(reshape2)
library(knockoff)
library(ggplot2)
library(rstan)

load("/Users/Patron/Documents/zinLDA research/count.Rdata")
load("/Users/Patron/Documents/zinLDA research/meta.RData")
norm_count <- count/rowSums(count)
col_means <- colMeans(norm_count>0)
indices <- which(col_means > 0.2)
sorted_indices <- indices[order(col_means[indices],decreasing = TRUE)]
dcount <- count[,sorted_indices]

X <- dcount
Y <- ifelse(meta$Group=="CRC",1,0)
```

We train the zinck model on $X$ using ADVI with an optimal number of clusters (20). We generate the knockoff matrix $\tilde{X}$ by plugging in the learnt  parameters into the generative model.

```{r zinck-fit, message=FALSE, warning=FALSE, results='hide'}
dlt <- rep(0,ncol(X)) ## Initializing deltas with the individual column sparsities

for(t in (1:ncol(X)))
{
  dlt[t] <- 1-mean(X[,t]>0)
  if(dlt[t]==0)
  {
    dlt[t] = dlt[t]+0.01
  }
  if (dlt[t]==1)
  {
    dlt[t] = dlt[t]-0.01
  }
  
}

zinLDA_stan_data <- list(
  K = 20,
  V = ncol(X),
  D = nrow(X),
  n = X,
  delta = dlt
)


zinck_code <- "data {
  int<lower=1> K; // num topics
  int<lower=1> V; // num words
  int<lower=0> D; // num docs
  int<lower=0> n[D, V]; // word counts for each doc

  // hyperparameters
  vector<lower=0, upper=1>[V] delta;
}

parameters {
  simplex[K] theta[D]; // topic mixtures
  vector<lower=0, upper=1>[V] zeta[K]; // zero-inflated betas
  vector<lower=0>[V] gamma1[K];
  vector<lower=0>[V] gamma2[K];
  vector<lower=0>[K] alpha;
}

transformed parameters {
  vector[V] beta[K];

  // Efficiently compute beta using vectorized operations
  for (k in 1:K) {
    vector[V] cum_log1m;
    cum_log1m[1:(V - 1)] = cumulative_sum(log1m(zeta[k, 1:(V - 1)]));
    cum_log1m[V] = 0;
    beta[k] = zeta[k] .* exp(cum_log1m);
    beta[k] = beta[k] / sum(beta[k]);
  }
}


model {
  for (k in 1:K) {
    alpha[k] ~ gamma(100,100);  // Change these hyperparameters as needed
  }
  for (d in 1:D) {
    theta[d] ~ dirichlet(alpha);
  }
  for (k in 1:K) {
    for (m in 1:V) {
        gamma1[k,m] ~ gamma(1,1);
        gamma2[k,m] ~ gamma(1,1);
    }
  }

  // Zero-inflated beta likelihood and data likelihood
  for (k in 1:K) {
    for (m in 1:V) {
      real lp_non_zero = bernoulli_lpmf(0 | delta[m]) + beta_lpdf(zeta[k, m] | gamma1[k, m], gamma2[k, m]);
      real lp_zero = bernoulli_lpmf(1 | delta[m]);
      target += log_sum_exp(lp_non_zero, lp_zero);
    }
  }

  // Compute the eta values and data likelihood more efficiently
  for (d in 1:D) {
    vector[V] eta = theta[d, 1] * beta[1];
    for (k in 2:K) {
      eta += theta[d, k] * beta[k];
    }
    eta = eta / sum(eta);
    n[d] ~ multinomial(eta);
  }
}
"

stan.model = stan_model(model_code = zinck_code)

set.seed(1)
fitCRC <- vb(stan.model, data=zinLDA_stan_data, algorithm="meanfield", iter=10000) ## Fitting the zinck model

theta <- fitCRC@sim[["est"]][["theta"]]
beta <- fitCRC@sim[["est"]][["beta"]]
X_tilde_CRC <- zinck::generateKnockoff(X,theta,beta,seed=1) ## Generating the knockoff copy
```

We move on to fit a tuned Random Forest model relating the augmented set of covariates with the outcome of interest $Y$.

```{r rf-model}
X_aug <- cbind(X,X_tilde_CRC) 
df_X <- as.data.frame(X_aug)
colnames(df_X) <- NULL
rownames(df_X) <- NULL
df_X$Y <- Y
set.seed(1)
model_rf <- randomForest(formula=as.factor(Y)~.,ntree=1000,mtry=28,importance=TRUE,data=as.matrix(df_X)) 
cf <- as.data.frame(importance(model_rf))[,3] ## Extracting the Mean Decrease in Impurities for each variable
W <- abs(cf[1:401])-abs(cf[402:802])
T <- knockoff.threshold(W,fdr = 0.1, offset = 1) ## This is the knockoff threshold
names_zinck <- colnames(X[,which(W>=T)])


```

Finally, we can visualize the importance of these selected species using the Feature Statistics obtained by contrasting the Random Forest importance scores of the original and the knockoff features.

```{r imp-plots}

names_zinck <- c("Ruminococcus torques species (1376)","Bacteroides caccae species (1382)","Clostridium symbiosum species (1475)","Hungatella hathewayi species (0882)","Unknown Ruminococcus species (6664)","Unknown Clostridiales species (6105)",  "Parvimonas micra species (1145)","Unknown Clostridium species (5413)","Gemella morbillorum species (4513)","Unknown Clostridiales species (6009)","Clostridium clostridioforme species (0979)", "Peptostreptococcus stomatis species (4614)","Solobacterium moorei species (0531)","Fusobacterium nucleatum species (0776)")


data.species <- data.frame(             ### Creating the data frame with Feature Importance Statistics
  impscores = sort(W[which(W>=T)], decreasing=FALSE) , 
  name = factor(names_zinck, levels = names_zinck),
  y = seq(length(names_zinck)) * 0.9
)

norm_count <- count/rowSums(count)
col_means <- colMeans(norm_count>0)
indices <- which(col_means > 0)
sorted_indices <- indices[order(col_means[indices],decreasing = TRUE)]
Xnorm <-norm_count[,sorted_indices]

# Calculate the column sums for cases and controls
case_sums <- colMeans(Xnorm[Y == 1, which(W>=T)])
control_sums <- colMeans(Xnorm[Y == 0, which(W>=T)])

# Determine colors based on the sum comparison
colors <- ifelse(case_sums > control_sums, "red", "blue")

# Create a data frame for plotting
data.species <- data.frame(
  impscores = sort(W[which(W >= T)], decreasing = FALSE),
  name = factor(names_zinck, levels = names_zinck),
  y = seq(length(names_zinck)) * 0.9,
  color = colors
)

plt.species <- ggplot(data.species) +
  geom_col(aes(impscores, name, fill = color), width = 0.6) +
  scale_fill_identity() +
  theme_bw() +
  ylab("Species") +
  xlab("Feature Statistic") +
  theme(
    axis.title.x = element_text(size = 22),
    axis.title.y = element_text(size = 22),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18, face = data.species$fontface) # Adjust fontface based on the condition
  )
print(plt.species)


```

