---
title: "Heatmaps"
author: "Soham Ghosh"
date: "2024-06-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Comparing the Heatmaps for Original and Knockoff Sample Taxa Matrices

We demonstrate the ability of zinck to capture the compositional and highly sparse nature of microbiome count data by comparing the heatmaps of the original sample taxa matrix $\mathbf{X}$ with its high quality knockoff copy, $\tilde{\mathbf{X}}$.

We consider a toy setting with $20$ samples and $30$ taxa with the library size between $400$ and $500$ for each sample, for a high zero-inflation level of $0.8$. We use the `simulateZINLDA()` function from the `zinLDA` package to generate the original sample taxa matrix. 

```{r heatmaps,message=FALSE, warning=FALSE,results='hide'}
library(zinck)
library(zinLDA)
library(ggplot2)
library(reshape2)
library(gridExtra)
library(cowplot)

set.seed(1) 
N.d=zinLDA::rdu(n=20,min=400,max=500) # Drawing random library sizes between 400, 500
sim_data = zinLDA::simulateZINLDA(D=20,V=30,N=N.d,K=5,Alpha=0.1,Pi=0.8,a=0.5,b=10)
X_original <- sim_data$sampleTaxaMatrix ## The original sample taxa count matrix


model_zinck <- fit.zinck(X_original, num_clusters=5, method="Gibbs", seed=1)
Theta <- model_zinck$theta
Beta <- model_zinck$beta

X_zinck <- generateKnockoff(X_original,Theta,Beta,seed=1)
rownames(X_zinck) <- rownames(X_original)

```

We will now visualize the heatmaps of the original matrix and its corresponding knockoff copy. The   function applies an arcsinh transformation to the data for normalization and better visualization of abundance patterns and zero inflation within the sample taxa matrix.

```{r heatmap-comparison, fig.width=15, fig.height=8}
heat1 <- draw_heatmap(X_original, "Original")
heat2 <- draw_heatmap(X_zinck, "Knockoff")
plot_grid(heat1, heat2, ncol = 2, align="v")
```


It is evident from the above heatmaps that the knockoff copy is almost indistinguishable from the original matrix! This underscores the fact that the knockoff copy preserves the underlying structure of the observed sample taxa count matrix. 
